---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---




## Loading and preprocessing the data
First check if everything is ok and load the data:

```{r preprocess} 
# make sure "activity.csv" is in the current directory
getwd()
list.files()
activity <- read.csv("activity.csv")

```



## What is mean total number of steps taken per day?
Following code computes the histogram of the total number of steps taken each day, ignoring NA values:

```{r histogram} 
total_steps_per_day <- activity %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps, na.rm=TRUE))

g <- with(total_steps_per_day, ggplot(total_steps_per_day, aes(total_steps))) + 
    geom_histogram() +
    ggtitle("Histogram of total steps") + 
    labs(x="Total steps", y="Count")

print(g)

```

Mean and median total number of steps taken per day:
```{r summary} 
stats <- summary(total_steps_per_day$total_steps)
mean <- format(stats[4], digits = 5)
median <- format(stats[3], digits = 5)
```
+ Mean = `r mean`
+ Median = `r median`



## What is the average daily activity pattern?

```{r timeseries}
ts_intervals <- activity %>%
    group_by(interval) %>%
    summarise(avg_steps = mean(steps, na.rm=TRUE))

g <- with(ts_intervals, ggplot(ts_intervals, aes(interval, avg_steps))) +
    geom_line()+
    ggtitle("Dayly pattern") + 
    labs(x="Interval", y="Steps dayly average ")

print(g)

```


## Imputing missing values
ust ignoring missing values could lead to biased conclusions. 
The following code evaluate the total number of missing values (NA)
in the dataset
```{r missings}

missing_values <-  nrow(activity[is.na(activity$steps), ])
total <- nrow(activity)
proportion <- format(missing_values*100 / total, digit = 2, nsmall = 2)

print(missing_values)
```

So that there are `r proportion`% missing values.

It could be interesting to use the mean of the corresponding interval
(which we just computed) instead of a missing value.
The following code just implements this strategy:

```{r fix_missings}

activity_fixed <- activity %>%
    merge(ts_intervals, by="interval") 



activity_fixed <- mutate(activity_fixed, steps =  ifelse(is.na(steps),
                                     avg_steps,
                                     steps))
activity_fixed <- select(activity_fixed, -avg_steps )

```
Let's view the corrected histogram of total number of steps per day:


```{r histogram_fixed} 
total_steps_per_day_fixed <- activity_fixed %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps, na.rm=TRUE))

g <- with(total_steps_per_day_fixed, ggplot(total_steps_per_day_fixed,
                                            aes(total_steps))) + 
    geom_histogram() +
    ggtitle("Histogram of total steps (fixing missing values)") + 
    labs(x="Total steps", y="Count")

print(g)

```

Corrected mean and median total number of steps taken per day:
```{r summary_fixed} 
stats <- summary(total_steps_per_day_fixed$total_steps)
mean <- format(stats[4], digits = 5)
median <- format(stats[3], digits = 5)
```
+ Mean = `r mean`
+ Median = `r median`



## Are there differences in activity patterns between weekdays and weekends?
It is expectable to see a variation of activity patterns between days and weekends.
Let's confirm this and characterise the variation, using the corrected dataset:

```{r weekdays_weekend}
# I18N issue : set locale to a fixed value
Sys.setlocale(locale="en_US")

activity_fixed <- activity_fixed %>%
    mutate(day_type = ifelse(weekdays(as.Date(date)) %in% c("Saturday", "Sunday"),
                             "weekend","weekday"))


#plot the time series side by side

ts_intervals_fixed <- activity_fixed %>%
    group_by(interval, day_type) %>%
    summarise(avg_steps = mean(steps, na.rm=TRUE))

g <- with(ts_intervals_fixed, ggplot(ts_intervals_fixed, aes(interval, avg_steps))) +
    geom_line()+
    facet_grid(. ~ day_type) +
    ggtitle("Dayly pattern") + 
    labs(x="Interval", y="Steps dayly average ")

print(g)


```
